Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry

### 2016-11-01 ms_pre_post.Rmd
Edit Results: Second subsection

Meanwhile, empirical studies of diverse species have found evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre- or post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre- or post-copulatory inbreeding avoidane despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These empirical studies demonstrate that diverse combinations of pre- and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no conceptual framework that predicts *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry can evolve. Such a framework would allow prediction of the form of mating system evolution across different systems that experience inbreeding depression and allow existing empirical studies to be used to test evolutionary predictions through comparative analyses.

@@

However, while @Duthie considered simultaneous evolution of polyandry and pre-copulatory inbreeding avoidance, they did not also allow simultaneous evolution of post-copulatory inbreeding avoidance, and assumed that polyandrous females had no post-copulatory control over fertilisation. If post-copulatory inbreeding avoidance could also evolve, polyandrous females could further bias paternity towards unrelated males within their set of chosen mates. This process might reduce the cost of polyandry stemming from accumulation of related mates, potentially facilitating evolution of polyandry to avoid inbreeding [@Zeh1997], and driving further consequent evolution of pre-copulatory or post-copulatory mate choice strategies. Therefore, while pre-copulatory inbreeding avoidance can facilitate evolution of polyandry under some restrictive conditions [@Duthie], the opportunity for post-copulatory choice might facilitate evolution of polyandry under a much wider range of conditions by allowing females to bias fertilisation toward non-kin. The long-term evolutionary dynamics of pre- and post-copulatory inbreeding avoidance then become unclear. Strong inbreeding depression might drive initial evolution of both pre- and post-copulatory inbreeding avoidance and associated polyandry. However, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance redundant, or vice versa, then only one strategy might be maintained in the long-term.

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models thereby isolate the invasion fitness of this single strategy. However, when both pre- and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

Ultimately, the absolute and relative frequencies of alleles underlying pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on resulting phenotypes (i.e., costs) and hence on the net costs versus benefits stemming from inbreeding avoidance. Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

Here we use individual-based modelling to set up *in silico* experiments, which test the hypothesis that, given strong inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance facilitates evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution such that only the less costly inbreeding avoidance strategy will be expressed in the long term. Finally, we test the hypothesis that evolution of one inbreeding avoidance strategy (i.e., pre- or post-copulatory) will be precluded if it is introduced into a population that is already fixed for the other inbreeding avoidance strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across different reproductive systems.

@@

@@

For simulations in which either pre- or post-copulatory inbreeding strategy allele values were fixed for random choice with respect to kinship ($M_{a}=0$ or $F_{a}=0$, respectively), the alternative costly inbreeding strategy evolved to inbreeding avoidance ($F_{a}<0$ or $M_{a}<0$; Figure 3). However, when the alternative inbreeding strategy could evolve and was not costly, the focal costly inbreeding strategy initially evolved towards inbreeding avoidance, but then evolved toward random choice with respect to kinship given increasingly negative allele values of the alternative strategy over generations (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ values were $-0.868$ in the last $10000$ generations (Figure 3A), meaning that an outbred focal female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. In contrast, grand mean $M_{a}$ allele values were $-0.288$ when post-copulatory inbreeding avoidance also evolved. More strongly, given fixed $M_{a}=0$, grand mean $F_{a}$ values were $-1.092$ in the last $10000$ generations (Figure 3C), meaning that an outbred focal female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, grand-mean $F_{a}$ values were $-0.070$ when pre-copulatory inbreeding strategy evolved, and polyandry allele values were negative. Overall, selection for a costly inbreeding strategy that was adaptive in isolation was greatly weakened given evolution of a less costly strategy.

@@

We have provided a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre- and post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre- or post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance, but not vice versa.

@@

In our model, a cost free inbreeding strategy precluded a more costly complementary inbreeding strategy from persisting in a focal population (Figure 2), meaning that long-term persistence of both pre- and post-copulatory inbreeding avoidance might not be expected in populations given cost asymmetry. Direct costs therefore strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through (3) loss of time or (4) depletion of energy. In our model, we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations.

@@

![Mean allele values underlying tendency for polyandry (red lines and shading), pre-copulatory inbreeding strategy (i.e., blue lines and shading), and post-copulatory inbreeding strategy (i.e., black lines and shading), as calculated across all individuals within a population over 40000 generations given strong inbreeding depression. Solid lines represent grand means over 40 replicate populations, and shading around solid lines show standard errors of allele values around grand means. Negative mean allele values reflect strategies of inbreeding avoidance or tendency for monandry, and positive values reflect strategies of inbreeding preference or tendency for polyandry. Panels in the left-hand column show simulations in which post-copulatory inbreeding strategy is fixed for random fertilisation (indicated by the black horizontal line), while panels in the right-hand column show evolution of both pre- and post-copulatory inbreeding avoidance. Costs of polyandry increase with rows from top to bottom. Dotted horizontal black lines in panels show values of zero on the y-axis.](figures/poly_costs.png)

![Relationship between polyandry allele values and phenotypes for an individual-based model in which pre- and post-copulatory inbreeding strategies evolve jointly with polyandry. Two simulations with identical starting conditions (default parameter values with no costs) are shown in panels A,B and C,D. Red lines in (A) and (C) show mean value of alleles underlying polyandry for a single replicate simulation over 40000 generations. Positive allele values contribute to polyandry, while negative values contribute to monandry (dotted line shows zero on the y-axis). In the last generation, mean allele value was below (A) or above (C) zero. Nevertheless, due to among-individual variation and the threshold nature of the polyandry phenotype, polyandry persists in the population in (A) and monandry persists in (C). The histograms in (B) and (D) show females' tendency for polyandry phenotypes in the last generation; white and grey shading indicates monandrous and polyandrous females, respectively.](figures/Pp_vals.png)

>>>>

Meanwhile, empirical studies of diverse species have found evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre-copulatoryor post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre-copulatoryor post-copulatory inbreeding avoidane despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These empirical studies demonstrate that diverse combinations of pre-copulatoryand post-copulatory inbreeding avoidance occur in nature. However, there is as yet no conceptual framework that predicts *a priori* what combinations should be favoured by selection when both pre-copulatoryand post-copulatory inbreeding avoidance, and polyandry can evolve. Such a framework would allow prediction of the form of mating system evolution across different systems that experience inbreeding depression and allow existing empirical studies to be used to test evolutionary predictions through comparative analyses.

@@

However, while @Duthie considered simultaneous evolution of polyandry and pre-copulatory inbreeding avoidance, they did not also allow simultaneous evolution of post-copulatory inbreeding avoidance, and assumed that polyandrous females had no post-copulatory control over fertilisation. If post-copulatory inbreeding avoidance could also evolve, polyandrous females could further bias paternity towards unrelated males within their set of chosen mates. This process might reduce the cost of polyandry stemming from accumulation of related mates, potentially facilitating evolution of polyandry to avoid inbreeding [@Zeh1997], and driving further consequent evolution of pre-copulatory or post-copulatory mate choice strategies. Therefore, while pre-copulatory inbreeding avoidance can facilitate evolution of polyandry under some restrictive conditions [@Duthie], the opportunity for post-copulatory choice might facilitate evolution of polyandry under a much wider range of conditions by allowing females to bias fertilisation toward non-kin. The long-term evolutionary dynamics of pre-copulatoryand post-copulatory inbreeding avoidance then become unclear. Strong inbreeding depression might drive initial evolution of both pre-copulatoryand post-copulatory inbreeding avoidance and associated polyandry. However, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance redundant, or vice versa, then only one strategy might be maintained in the long-term.

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models thereby isolate the invasion fitness of this single strategy. However, when both pre-copulatoryand post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatoryand post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on resulting phenotypes (i.e., costs) and hence on the net costs versus benefits stemming from inbreeding avoidance. Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre-copulatoryand post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

Here we use individual-based modelling to set up *in silico* experiments, which test the hypothesis that, given strong inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance facilitates evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution such that only the less costly inbreeding avoidance strategy will be expressed in the long term. Finally, we test the hypothesis that evolution of one inbreeding avoidance strategy (i.e., pre-copulatoryor post-copulatory) will be precluded if it is introduced into a population that is already fixed for the other inbreeding avoidance strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across different reproductive systems.

@@

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero and could not evolve, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

<!--- Co-author wants to talk about polyandry and stuff: But these figures don’t look very symmetrical – there’s stuff going on in Fig 3 that isn’t really noted in the text – should we do a bit more with this? The most obvious difference is that polyandry evolves in B but not D - i.e. when pre-cop IA is costly but not when post-cop IA is costly. Wouldn’t it be helpful to point this out and explain it? NB, presumably polyandry is cost-free in all these simulations? Might it be worth re-iterating that too? --->

@@

@@

We have provided a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre-copulatoryand post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre-copulatoryand post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre-copulatoryor post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance, but not vice versa.

@@

In our model, a cost free inbreeding strategy precluded a more costly complementary inbreeding strategy from persisting in a focal population (Figure 2), meaning that long-term persistence of both pre-copulatoryand post-copulatory inbreeding avoidance might not be expected in populations given cost asymmetry. Direct costs therefore strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through (3) loss of time or (4) depletion of energy. In our model, we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations.

@@

![Mean allele values underlying tendency for polyandry (red lines and shading), pre-copulatory inbreeding strategy (i.e., blue lines and shading), and post-copulatory inbreeding strategy (i.e., black lines and shading), as calculated across all individuals within a population over 40000 generations given strong inbreeding depression. Solid lines represent grand means over 40 replicate populations, and shading around solid lines show standard errors of allele values around grand means. Negative mean allele values reflect strategies of inbreeding avoidance or tendency for monandry, and positive values reflect strategies of inbreeding preference or tendency for polyandry. Panels in the left-hand column show simulations in which post-copulatory inbreeding strategy is fixed for random fertilisation (indicated by the black horizontal line), while panels in the right-hand column show evolution of both pre-copulatoryand post-copulatory inbreeding avoidance. Costs of polyandry increase with rows from top to bottom. Dotted horizontal black lines in panels show values of zero on the y-axis.](figures/poly_costs.png)

![Relationship between polyandry allele values and phenotypes for an individual-based model in which pre-copulatoryand post-copulatory inbreeding strategies evolve jointly with polyandry. Two simulations with identical starting conditions (default parameter values with no costs) are shown in panels A,B and C,D. Red lines in (A) and (C) show mean value of alleles underlying polyandry for a single replicate simulation over 40000 generations. Positive allele values contribute to polyandry, while negative values contribute to monandry (dotted line shows zero on the y-axis). In the last generation, mean allele value was below (A) or above (C) zero. Nevertheless, due to among-individual variation and the threshold nature of the polyandry phenotype, polyandry persists in the population in (A) and monandry persists in (C). The histograms in (B) and (D) show females' tendency for polyandry phenotypes in the last generation; white and grey shading indicates monandrous and polyandrous females, respectively.](figures/Pp_vals.png)
